---
title: "Merge samples"
author: "Mechthild Lütge"
date: "14 May 2020"
output:
  html_document:
    self_contained: no
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
set.seed(1000)
```

## load packages
```{r libs}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(tidyverse)
  library(Seurat)
  library(magrittr)
  library(dplyr)
  library(purrr)
  library(ggplot2)
  library(here)
  library(runSeurat3)
  library(ggsci)
  library(ggpubr)
  library(pheatmap)
  library(viridis)
  library(sctransform)
})

```


## load data 
```{r set dir}
basedir <- here()
seurat <- readRDS(file = paste0(basedir, 
                              "/data/humanHeartsPlusGraz_intPatients_merged", 
                              "_seurat.rds"))
Idents(seurat) <- seurat$seurat_clusters
```

## exclude two patients with very low nuclei
```{r exclude two patients}

## two patients with <250 nuclei --> exclude from downstream analysis
table(seurat$ID)
seurat <- subset(seurat, ID %in% c("GZ19", "SG30"), invert=T)
table(seurat$ID)


```


## assign labels
```{r assign labels}

seurat$label <- "other"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("2","9","7"))] <- "Endothelial"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("10"))] <- "EndoEC"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("12"))] <- "LEC"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("4"))] <- "Tcell"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("3"))] <- "Cardiomyocyte"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("0"))] <- "Fibroblast"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("1"))] <- "Perivascular"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("8"))] <- "SMC"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("5"))] <- "resMacrophage"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("6"))] <- "infMacrophage"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("11"))] <- "NeuralCells"
seurat$label[which(seurat$integrated_snn_res.0.4 %in% c("13"))] <- "Adipocytes"

```



## color vectors
```{r colVec}

colPal <- c(pal_igv()(12),
            pal_aaas()(10))[1:length(levels(seurat))]
colTec <- pal_jama()(length(unique(seurat$technique)))
colSmp <- c(pal_uchicago()(9), pal_npg()(10), pal_aaas()(10), 
            pal_jama()(7))[1:length(unique(seurat$dataset))]
colCond <- pal_npg()(length(unique(seurat$cond)))
colID <- c(pal_jco()(10), pal_npg()(10), pal_futurama()(10),
           pal_d3()(10))[1:length(unique(seurat$ID))]
colOrig <- pal_aaas()(length(unique(seurat$origin)))
colIso <- pal_nejm()(length(unique(seurat$isolation)))
colProc <- pal_aaas()(length(unique(seurat$processing)))
colLab <- c(pal_futurama()(8), pal_uchicago()(6))[1:length(unique(seurat$label))]

names(colPal) <- levels(seurat)
names(colTec) <- unique(seurat$technique)
names(colSmp) <- unique(seurat$dataset)
names(colCond) <- unique(seurat$cond)
names(colID) <- unique(seurat$ID)
names(colOrig) <- unique(seurat$origin)
names(colIso) <- unique(seurat$isolation)
names(colProc) <- unique(seurat$processing)
names(colLab) <- unique(seurat$label)

```


## vis data {.tabset}
### clusters
```{r vis cluster}
DimPlot(seurat, reduction = "umap", cols=colPal)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

DimPlot(seurat, reduction = "umap", cols=colPal,
        shuffle = T)+
  theme_void()

```

### label
```{r vis label}
DimPlot(seurat, reduction = "umap",  group.by = "label", cols=colLab)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

DimPlot(seurat, reduction = "umap", group.by = "label", cols=colLab,
        shuffle = T)+
  theme_void()
```


### technique
```{r vis tecnique}

DimPlot(seurat, reduction = "umap", group.by = "technique", cols=colTec)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

```


### Sample
```{r vis sample, fig.height=4, fig.width=12}

DimPlot(seurat, reduction = "umap", group.by = "dataset", cols=colSmp)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### ID
```{r vis ID}

DimPlot(seurat, reduction = "umap", group.by = "ID", cols=colID, shuffle = T)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

DimPlot(seurat, reduction = "umap", group.by = "ID", cols=colID,
        shuffle = T)+
  theme_void()

```


### Origin
```{r vis origin}

DimPlot(seurat, reduction = "umap", group.by = "origin", cols=colOrig,
        shuffle = T)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

DimPlot(seurat, reduction = "umap", group.by = "origin", cols=colOrig,
        shuffle = T)+
  theme_void()
```


### isolation
```{r vis isolation}

DimPlot(seurat, reduction = "umap", group.by = "isolation", cols=colIso)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### cond
```{r vis cond, fig.height=4, fig.width=5}

DimPlot(seurat, reduction = "umap", group.by = "cond", cols=colCond)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### processing
```{r vis proc}

DimPlot(seurat, reduction = "umap", group.by = "processing", cols=colProc)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

## cnt tables {.tabset}
### per patient
```{r cnt tables per pat, fig.height=6, fig.width=8}

## total cells per patient
knitr::kable(table(seurat$ID))

## celltype per patient counts
knitr::kable(table(seurat$label, seurat$ID))

## celltype percentages per patient
datLab <- data.frame(table(seurat$label, seurat$ID))
colnames(datLab) <- c("label", "ID", "cnt")
datPat <- data.frame(table(seurat$ID))
colnames(datPat) <- c("ID", "total")
datFrac <- datLab %>% left_join(., datPat, by="ID") %>% 
  mutate(percentage = cnt*100/total)
knitr::kable(datFrac)

ordVec <- datFrac %>% dplyr::filter(label=="Tcell") %>% 
  arrange(., percentage)

ggbarplot(datFrac, x="ID", y="percentage",
          fill = "label",
          palette = colLab,
          order= ordVec$ID) +
  rotate_x_text(angle = 90)

```

### per cond
```{r cnt tables per cond, fig.height=6, fig.width=6}

## total cells per cond
knitr::kable(table(seurat$cond))

## celltype per cond counts
knitr::kable(table(seurat$label, seurat$cond))

## celltype percentages per cond
datLab <- data.frame(table(seurat$label, seurat$cond))
colnames(datLab) <- c("label", "cond", "cnt")
datPat <- data.frame(table(seurat$cond))
colnames(datPat) <- c("cond", "total")
datFrac <- datLab %>% left_join(., datPat, by="cond") %>% 
  mutate(percentage = cnt*100/total)
knitr::kable(datFrac)

ggbarplot(datFrac, x="cond", y="percentage",
          fill = "label",
          palette = colLab) +
  rotate_x_text(angle = 90)

```

## assign groups based on T cell fraction
```{r assign groups}

TcellGrp <- read_tsv(paste0(basedir, "/data/assignTcellGrp.txt"))
IDtoTcell <- data.frame(ID=seurat$ID) %>% left_join(., TcellGrp, by="ID")
seurat$TcellGrp <- IDtoTcell$TcellGrp
table(seurat$TcellGrp)
table(seurat$TcellGrp, seurat$ID)
```


## vis marker for celltype assignment

```{r avg heatmap function, eval=T, include=FALSE}

avgHeatmap <- function(seurat, selGenes, colVecIdent, colVecCond=NULL,
                       ordVec=NULL, gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                       cr=FALSE, condCol=FALSE){
  
  ## format gene names (depends on how gene list is inputed)
  selGenes <- selGenes %>% 
      mutate(labelNam = (str_split(gene, '\\.', simplify = T)[,2]))
  selGenes <- selGenes$labelNam
  
  ## assay data
  clusterAssigned <- as.data.frame(Idents(seurat)) %>%
  dplyr::mutate(cell=rownames(.))
  colnames(clusterAssigned)[1] <- "ident"
  seuratDat <- GetAssayData(seurat)
  
  ## genes of interest
  genes <- data.frame(gene=rownames(seurat)) %>% 
    mutate(geneID=gsub("^.*\\.", "", gene)) %>% filter(geneID %in% selGenes)

  ## matrix with averaged cnts per ident
  logNormExpres <- as.data.frame(t(as.matrix(
    seuratDat[which(rownames(seuratDat) %in% genes$gene),])))
  logNormExpres <- logNormExpres %>% dplyr::mutate(cell=rownames(.)) %>%
    dplyr::left_join(.,clusterAssigned, by=c("cell")) %>%
    dplyr::select(-cell) %>% dplyr::group_by(ident) %>%
    dplyr::summarise_all(mean)
  logNormExpresMa <- logNormExpres %>% dplyr::select(-ident) %>% as.matrix()
  rownames(logNormExpresMa) <- logNormExpres$ident
  logNormExpresMa <- t(logNormExpresMa)
  rownames(logNormExpresMa) <- gsub("^.*?\\.","",rownames(logNormExpresMa))
  
  ## remove genes if they are all the same in all groups
  ind <- apply(logNormExpresMa, 1, sd) == 0
  logNormExpresMa <- logNormExpresMa[!ind,]
  genes <- genes[!ind,]

  ## color columns according to cluster
  annotation_col <- as.data.frame(gsub("(^.*?_)","",
                                       colnames(logNormExpresMa)))%>%
    dplyr::mutate(celltype=gsub("(_.*$)","",colnames(logNormExpresMa)))
  colnames(annotation_col)[1] <- "col1"
  annotation_col <- annotation_col %>%
    dplyr::mutate(cond = gsub("(^[0-9]_?)","",col1)) %>%
    dplyr::select(cond, celltype)
  rownames(annotation_col) <- colnames(logNormExpresMa) 

  ann_colors = list(
      cond = colVecCond,
      celltype=colVecIdent)
  if(is.null(ann_colors$cond)){
    annotation_col$cond <- NULL
  }
  
  ## adjust order
  logNormExpresMa <- logNormExpresMa[selGenes,]
  if(is.null(ordVec)){
    ordVec <- levels(seurat)
  }
  logNormExpresMa <- logNormExpresMa[,ordVec]

  ## scaled row-wise
  pheatmap(logNormExpresMa, scale="row" ,treeheight_row = 0, cluster_rows = cr, 
         cluster_cols = cc,
         color = colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50),
         annotation_col = annotation_col, cellwidth=15, cellheight=10,
         annotation_colors = ann_colors, gaps_row = gapVecR, gaps_col = gapVecC)
}

```

### heatmap
```{r vis sel marker, fig.height=10, fig.width=8}

genes <- data.frame(gene=rownames(seurat)) %>% 
    mutate(geneID=gsub("^.*\\.", "", gene)) 

selGenesAll <- read_tsv(file = paste0(basedir,
                                      "/data/markerLabels.txt")) %>% 
  left_join(., genes, by = "geneID")

Idents(seurat) <- seurat$seurat_clusters
pOut <- avgHeatmap(seurat = seurat, selGenes = selGenesAll,
                  colVecIdent = colPal, 
                  ordVec=levels(seurat),
                  gapVecR=NULL, gapVecC=NULL,cc=T,
                  cr=F, condCol=F)

Idents(seurat) <- seurat$label
pOut <- avgHeatmap(seurat = seurat, selGenes = selGenesAll,
                  colVecIdent = colLab, 
                  ordVec=levels(seurat),
                  gapVecR=NULL, gapVecC=NULL,cc=T,
                  cr=F, condCol=F)

```

### Dotplot
```{r dotplot, fig.height=9, fig.width=7}

DotPlot(seurat, assay="RNA", features = selGenesAll$gene, scale =T,
        cluster.idents = T) +
  scale_color_viridis_c() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(breaks=selGenesAll$gene, labels=selGenesAll$geneID) +
  xlab("") + ylab("")

Idents(seurat) <- seurat$seurat_clusters
DotPlot(seurat, assay="RNA", features = selGenesAll$gene, scale =T,
        cluster.idents = T) +
  scale_color_viridis_c() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(breaks=selGenesAll$gene, labels=selGenesAll$geneID) +
  xlab("") + ylab("")

```

## violin plots selected marker
```{r vlnplt sel marker}

genesDat <- data.frame(EnsID=rownames(seurat)) %>% 
  mutate(gene=gsub(".*\\.", "", EnsID))
selGenes <- data.frame(gene=c("CD2", "TNNT2", "PECAM1", "NRG1", "PROX1",
                              "PDGFRA", "RGS5", "MYH11", "C1QA", "NRXN1",
                              "PLIN1", "BMP4", "BMPR1A", "BMPR2")) %>% 
  left_join(., genesDat, by="gene")

pList <- sapply(selGenes$EnsID, function(x){
  p <- VlnPlot(object = seurat, features = x,
               group.by = "label",
               cols = colLab, pt.size = 0
               ) + 
    theme(legend.position = "none")
    
  plot(p)
})

pList <- sapply(selGenes$EnsID, function(x){
  p <- VlnPlot(object = seurat, features = x,
               group.by = "label",
               cols = colLab, pt.size = 0.3
               ) + 
    theme(legend.position = "none")
    
  plot(p)
})


```

## Feature plots BMPs
```{r feautreplots BMPs}

## list with all gene names for mapping of EnsIDs
genesDat <- data.frame(EnsID=rownames(seurat)) %>% 
  mutate(gene=gsub(".*\\.", "", EnsID))

## selected genes to plot
selGenes <- data.frame(gene=c("BMP2", "BMP4", "BMPR1A", "BMPR2")) %>% 
  left_join(., genesDat, by="gene")

## plotting loop order=F
pList <- sapply(selGenes$EnsID, function(x){
p <- FeaturePlot(seurat, reduction = "umap", 
            features = x,
            cols=c("lightgrey", "darkred"),
            order = F)+
  theme(legend.position="right")
  plot(p)
})

## plotting loop order=T
pList <- sapply(selGenes$EnsID, function(x){
p <- FeaturePlot(seurat, reduction = "umap", 
            features = x,
            cols=c("lightgrey", "darkred"),
            order = T)+
  theme(legend.position="right")
  plot(p)
})
```


## average counts BMPs in Fibroblasts
```{r average cnts BMPs}

seuratSub <- subset(seurat, label=="Fibroblast")

## assay data
clusterAssigned <- as.data.frame(seuratSub$ID) %>%
  dplyr::mutate(cell=rownames(.))
colnames(clusterAssigned)[1] <- "ident"
seuratDat <- GetAssayData(seuratSub)
  
  ## genes of interest
genes <- data.frame(gene=rownames(seuratSub)) %>% 
  mutate(geneID=gsub("^.*\\.", "", gene)) %>% filter(geneID %in% selGenes$gene)

  ## matrix with averaged cnts per ident
logNormExpres <- as.data.frame(t(as.matrix(
  seuratDat[which(rownames(seuratDat) %in% genes$gene),])))
logNormExpres <- logNormExpres %>% dplyr::mutate(cell=rownames(.)) %>%
  dplyr::left_join(.,clusterAssigned, by=c("cell")) %>%
  dplyr::select(-cell) %>% dplyr::group_by(ident) %>%
  dplyr::summarise_all(mean)

write.table(logNormExpres, 
            file=paste0(basedir, "/data/BmpCntsFibroblastsPerPatient.txt"),
            row.names = F, col.names = T, sep = "\t", quote = F)
```


## save seurat object
```{r save}
saveRDS(seurat, file = paste0(basedir, 
                              "/data/humanHeartsPlusGraz_intPatients_merged", 
                              "labeled_seurat.rds"))
```


## session info
```{r session info}
sessionInfo()
date()
```



