---
title: "cluster samples"
author: "Mechthild Lütge"
date: "14 May 2020"
output:
  html_document:
    self_contained: no
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
set.seed(1000)
```

## load packages
```{r libs}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(tidyverse)
  library(Seurat)
  library(magrittr)
  library(dplyr)
  library(purrr)
  library(ggplot2)
  library(here)
  library(runSeurat3)
  library(ggsci)
  library(ggpubr)
  library(pheatmap)
  library(viridis)
  library(scran)
  library(edgeR)
  library(muscat)
})

```


## load data 
```{r set dir}
basedir <- here()
seurat <- readRDS(file = paste0(basedir, 
                              "/data/humanHeartsPlusGraz_intPatients_merged", 
                              "labeled_seurat.rds"))
Idents(seurat) <- seurat$ID
```


## color vectors
```{r colVec}

colPal <- c(pal_igv()(12),
            pal_aaas()(10))[1:length(levels(seurat))]
colTec <- pal_jama()(length(unique(seurat$technique)))
colSmp <- c(pal_uchicago()(9), pal_npg()(10), pal_aaas()(10), 
            pal_jama()(7))[1:length(unique(seurat$dataset))]
colCond <- pal_npg()(length(unique(seurat$cond)))
colID <- c(pal_jco()(10), pal_npg()(10), pal_futurama()(10),
           pal_d3()(10))[1:length(unique(seurat$ID))]
colOrig <- pal_aaas()(length(unique(seurat$origin)))
colIso <- pal_nejm()(length(unique(seurat$isolation)))
colTG <- pal_aaas()(length(unique(seurat$TcellGrp)))
colLab <- pal_futurama()(length(unique(seurat$label)))

names(colPal) <- levels(seurat)
names(colTec) <- unique(seurat$technique)
names(colSmp) <- unique(seurat$dataset)
names(colCond) <- unique(seurat$cond)
names(colID) <- unique(seurat$ID)
names(colOrig) <- unique(seurat$origin)
names(colIso) <- unique(seurat$isolation)
names(colTG) <- unique(seurat$TcellGrp)
names(colLab) <- unique(seurat$label)

```


## vis data {.tabset}

### label
```{r vis label}
DimPlot(seurat, reduction = "umap",  group.by = "label", cols=colLab)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### technique
```{r vis tecnique}

DimPlot(seurat, reduction = "umap", group.by = "technique", cols=colTec)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

```


### Sample
```{r vis sample, fig.height=4, fig.width=12}

DimPlot(seurat, reduction = "umap", group.by = "dataset", cols=colSmp)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### ID
```{r vis ID}

DimPlot(seurat, reduction = "umap", group.by = "ID", cols=colID)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### Origin
```{r vis origin}

DimPlot(seurat, reduction = "umap", group.by = "origin", cols=colOrig)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### cond
```{r vis cond, fig.height=4, fig.width=5}

DimPlot(seurat, reduction = "umap", group.by = "cond", cols=colCond)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### T cell grp
```{r vis Tcell grp}

DimPlot(seurat, reduction = "umap", group.by = "TcellGrp", cols=colTG)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

## cluster patients without downsampling{.tabset}
### vis and cluster based on celltype abundance
```{r vis patient celltype abundance in heatmap}
## remove "other" cells
seurat <- subset(seurat, label=="other", invert=T)

## prepare seurat object
DefaultAssay(seurat) <- "RNA"
seurat <- FindVariableFeatures(seurat)
seurat <- ScaleData(seurat)
seurat <- RunPCA(seurat)

sce <- as.SingleCellExperiment(seurat)
by.label <- table(colLabels(sce), sce$ID)
patGrp <- data.frame(unique(cbind(sce$ID, sce$TcellGrp)))
colnames(patGrp) <- c("ID", "TcellGrp")
patGrpCol <- patGrp %>% dplyr::select(TcellGrp)
rownames(patGrpCol) <- patGrp$ID
pheatmap::pheatmap(log2(by.label+1), color=viridis::viridis(101),
                   annotation_col = patGrpCol,
                   annotation_colors = list(TcellGrp=colTG))
```

### perform pseudo-bulking
```{r pseudobulking}

altExp(sce) <- NULL
summed <- aggregateAcrossCells(sce,
    id=colData(sce)[,c("ID")])
summed$ncells

y <- DGEList(counts(summed), samples=colData(summed))
keep <- filterByExpr(y, group=summed$TcellGrp)
y <- y[keep,]
summary(keep)
y <- calcNormFactors(y)

for (i in seq_len(ncol(y))) {
    plotMD(y, column=i)
}

plotMDS(cpm(y, log=TRUE), 
    col=ifelse(y$samples$TcellGrp == "Ctrl", colTG["Ctrl"],
               ifelse(y$samples$TcellGrp == "TcellLow",
                      colTG["TcellLow"], colTG["TcellHigh"])))

```

### muscat
```{r muscat}
sce <- prepSCE(sce, 
    kid = "TcellGrp", 
    gid = "TcellGrp",  
    sid = "ID",   
    drop = TRUE)

nk <- length(kids <- levels(sce$cluster_id))
ns <- length(sids <- levels(sce$sample_id))
names(kids) <- kids; names(sids) <- sids

pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id"))
# one sheet per subpopulation
assayNames(pb)

pbMDS(pb)

```


## cluster patients after downsampling{.tabset}
### vis and cluster based on celltype abundance
```{r vis patient celltype abundance in heatmap ds}

table(seurat$ID)
min(table(seurat$ID))
max(table(seurat$ID))
Idents(seurat) <- seurat$ID
seuratSub <- subset(seurat, downsample = 500)


sce <- as.SingleCellExperiment(seuratSub)
by.label <- table(colLabels(sce), sce$ID)
patGrp <- data.frame(unique(cbind(sce$ID, sce$TcellGrp)))
colnames(patGrp) <- c("ID", "TcellGrp")
patGrpCol <- patGrp %>% dplyr::select(TcellGrp)
rownames(patGrpCol) <- patGrp$ID
pheatmap::pheatmap(log2(by.label+1), color=viridis::viridis(101),
                   annotation_col = patGrpCol,
                   annotation_colors = list(TcellGrp=colTG))
```


### perform pseudo-bulking
```{r pseudobulking ds}

altExp(sce) <- NULL
summed <- aggregateAcrossCells(sce,
    id=colData(sce)[,c("ID")])
summed$ncells

y <- DGEList(counts(summed), samples=colData(summed))
keep <- filterByExpr(y, group=summed$TcellGrp)
y <- y[keep,]
summary(keep)
y <- calcNormFactors(y)

for (i in seq_len(ncol(y))) {
    plotMD(y, column=i)
}

plotMDS(cpm(y, log=TRUE), 
    col=ifelse(y$samples$TcellGrp == "Ctrl", colTG["Ctrl"],
               ifelse(y$samples$TcellGrp == "TcellLow",
                      colTG["TcellLow"], colTG["TcellHigh"])))

```

### muscat
```{r muscat ds}
sce <- prepSCE(sce, 
    kid = "TcellGrp", 
    gid = "TcellGrp",  
    sid = "ID",   
    drop = TRUE)

nk <- length(kids <- levels(sce$cluster_id))
ns <- length(sids <- levels(sce$sample_id))
names(kids) <- kids; names(sids) <- sids

pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id"))
# one sheet per subpopulation
assayNames(pb)

pbMDS(pb)

```


## cluster patients after downsampling withoout highly active pat{.tabset}
### vis and cluster based on celltype abundance
```{r vis patient celltype abundance in heatmap ds wo hiAct pat}

seuratSub <- subset(seuratSub, ID %in% c("SG29", "SG33"), invert=T)

sce <- as.SingleCellExperiment(seuratSub)
by.label <- table(colLabels(sce), sce$ID)
patGrp <- data.frame(unique(cbind(sce$ID, sce$TcellGrp)))
colnames(patGrp) <- c("ID", "TcellGrp")
patGrpCol <- patGrp %>% dplyr::select(TcellGrp)
rownames(patGrpCol) <- patGrp$ID
pheatmap::pheatmap(log2(by.label+1), color=viridis::viridis(101),
                   annotation_col = patGrpCol,
                   annotation_colors = list(TcellGrp=colTG))
```


### perform pseudo-bulking
```{r pseudobulking ds wo hiAct pat}

altExp(sce) <- NULL
summed <- aggregateAcrossCells(sce,
    id=colData(sce)[,c("ID")])
summed$ncells

y <- DGEList(counts(summed), samples=colData(summed))
keep <- filterByExpr(y, group=summed$TcellGrp)
y <- y[keep,]
summary(keep)
y <- calcNormFactors(y)

for (i in seq_len(ncol(y))) {
    plotMD(y, column=i)
}

plotMDS(cpm(y, log=TRUE), 
    col=ifelse(y$samples$TcellGrp == "Ctrl", colTG["Ctrl"],
               ifelse(y$samples$TcellGrp == "TcellLow",
                      colTG["TcellLow"], colTG["TcellHigh"])))

```

### muscat
```{r muscat ds wo hiAct pat}
sce <- prepSCE(sce, 
    kid = "TcellGrp", 
    gid = "TcellGrp",  
    sid = "ID",   
    drop = TRUE)

nk <- length(kids <- levels(sce$cluster_id))
ns <- length(sids <- levels(sce$sample_id))
names(kids) <- kids; names(sids) <- sids

pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id"))
# one sheet per subpopulation
assayNames(pb)

pbMDS(pb)

```



## session info
```{r session info}
sessionInfo()
date()
```



