---
title: "Merge samples"
author: "Mechthild Lütge"
date: "14 May 2020"
output:
  html_document:
    self_contained: no
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
set.seed(1000)
```

## load packages
```{r libs}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(tidyverse)
  library(Seurat)
  library(magrittr)
  library(dplyr)
  library(purrr)
  library(ggplot2)
  library(here)
  library(runSeurat3)
  library(ggsci)
  library(ggpubr)
  library(pheatmap)
  library(viridis)
  library(sctransform)
  library(fgsea)
  library(grid)
  library(gridExtra)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(DOSE)
  library(enrichplot)
  library(msigdbr)

})

```


```{r avg heatmap function, eval=T, include=FALSE}

avgHeatmap <- function(seurat, selGenes, colVecIdent, colVecCond=NULL,
                       ordVec=NULL, gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                       cr=FALSE, condCol=FALSE){
  
  ## format gene names (depends on how gene list is inputed)
  selGenes <- selGenes %>% 
      mutate(labelNam = (str_split(gene, '\\.', simplify = T)[,2]))
  selGenes <- selGenes$labelNam
  
  ## assay data
  clusterAssigned <- as.data.frame(Idents(seurat)) %>%
  dplyr::mutate(cell=rownames(.))
  colnames(clusterAssigned)[1] <- "ident"
  seuratDat <- GetAssayData(seurat)
  
  ## genes of interest
  genes <- data.frame(gene=rownames(seurat)) %>% 
    mutate(geneID=gsub("^.*\\.", "", gene)) %>% filter(geneID %in% selGenes)

  ## matrix with averaged cnts per ident
  logNormExpres <- as.data.frame(t(as.matrix(
    seuratDat[which(rownames(seuratDat) %in% genes$gene),])))
  logNormExpres <- logNormExpres %>% dplyr::mutate(cell=rownames(.)) %>%
    dplyr::left_join(.,clusterAssigned, by=c("cell")) %>%
    dplyr::select(-cell) %>% dplyr::group_by(ident) %>%
    dplyr::summarise_all(mean)
  logNormExpresMa <- logNormExpres %>% dplyr::select(-ident) %>% as.matrix()
  rownames(logNormExpresMa) <- logNormExpres$ident
  logNormExpresMa <- t(logNormExpresMa)
  rownames(logNormExpresMa) <- gsub("^.*?\\.","",rownames(logNormExpresMa))
  
  ## remove genes if they are all the same in all groups
  ind <- apply(logNormExpresMa, 1, sd) == 0
  logNormExpresMa <- logNormExpresMa[!ind,]
  genes <- genes[!ind,]

  ## color columns according to cluster
  annotation_col <- as.data.frame(gsub("(^.*?_)","",
                                       colnames(logNormExpresMa)))%>%
    dplyr::mutate(celltype=gsub("(_.*$)","",colnames(logNormExpresMa)))
  colnames(annotation_col)[1] <- "col1"
  annotation_col <- annotation_col %>%
    dplyr::mutate(cond = gsub("(^[0-9]_?)","",col1)) %>%
    dplyr::select(cond, celltype)
  rownames(annotation_col) <- colnames(logNormExpresMa) 

  ann_colors = list(
      cond = colVecCond,
      celltype=colVecIdent)
  if(is.null(ann_colors$cond)){
    annotation_col$cond <- NULL
  }
  
  ## adjust order
  logNormExpresMa <- logNormExpresMa[selGenes,]
  if(is.null(ordVec)){
    ordVec <- levels(seurat)
  }
  logNormExpresMa <- logNormExpresMa[,ordVec]

  ## scaled row-wise
  pheatmap(logNormExpresMa, scale="row" ,treeheight_row = 0, cluster_rows = cr, 
         cluster_cols = cc,
         color = colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50),
         annotation_col = annotation_col, cellwidth=15, cellheight=10,
         annotation_colors = ann_colors, gaps_row = gapVecR, gaps_col = gapVecC)
}

```


## sign plot funct
```{r funct sign plot}
## random plotting order
shuf <- function(df){
  return(df[sample(1:dim(df)[1], dim(df)[1]),])
}

## adapted from CellMixS
visGroup_adapt <- function (sce,group,dim_red = "TSNE",col_group=pal_nejm()(8)) 
{
    if (!is(sce, "SingleCellExperiment")) {
        stop("Error:'sce' must be a 'SingleCellExperiment' object.")
    }
    if (!group %in% names(colData(sce))) {
        stop("Error: 'group' variable must be in 'colData(sce)'")
    }
    cell_names <- colnames(sce)
    if (!dim_red %in% "TSNE") {
        if (!dim_red %in% reducedDimNames(sce)) {
            stop("Please provide a dim_red method listed in reducedDims of sce")
        }
        red_dim <- as.data.frame(reducedDim(sce, dim_red))
    }
    else {
        if (!"TSNE" %in% reducedDimNames(sce)) {
            if ("logcounts" %in% names(assays(sce))) {
                sce <- runTSNE(sce)
            }
            else {
                sce <- runTSNE(sce, exprs_values = "counts")
            }
        }
        red_dim <- as.data.frame(reducedDim(sce, "TSNE"))
    }
    colnames(red_dim) <- c("red_dim1", "red_dim2")
    df <- data.frame(sample_id = cell_names, group_var = colData(sce)[, 
        group], red_Dim1 = red_dim$red_dim1, red_Dim2 = red_dim$red_dim2)
    t <- ggplot(shuf(df), aes_string(x = "red_Dim1", y = "red_Dim2")) + 
        xlab(paste0(dim_red, "_1")) + ylab(paste0(dim_red, "_2")) + 
        theme_void() + theme(aspect.ratio = 1,
                             panel.grid.minor = element_blank(), 
        panel.grid.major = element_line(color = "grey", size = 0.3))
    t_group <- t + geom_point(size = 1, alpha = 0.7,
                              aes_string(color = "group_var")) + 
        guides(color = guide_legend(override.aes = list(size = 1), 
            title = group)) + ggtitle(group)
    if (is.numeric(df$group_var)) {
        t_group <- t_group + scale_color_viridis(option = "D")
    }
    else {
        t_group <- t_group + scale_color_manual(values = col_group)
    }
    t_group
}



```


## integrate data
```{r set dir}
basedir <- here()
seurat <- readRDS(file = paste0(basedir, 
                              "/data/humanHeartsPlusGraz_intPatients_", 
                              "merged_Subset_seurat.rds"))

```



## color vectors
```{r colVec}

colPal <- pal_igv()(length(levels(seurat)))
colTec <- pal_jama()(length(unique(seurat$technique)))
colSmp <- c(pal_uchicago()(8), pal_npg()(8), pal_aaas()(10))[1:length(unique(seurat$dataset))]
colCond <- pal_npg()(length(unique(seurat$cond2)))
colID <- c(pal_jco()(10), pal_npg()(10))[1:length(unique(seurat$ID))]
colOrig <- pal_aaas()(length(unique(seurat$origin)))
colIso <- pal_nejm()(length(unique(seurat$isolation)))
colProc <- pal_aaas()(length(unique(seurat$processing)))
colLab <- pal_futurama()(length(unique(seurat$label)))

names(colPal) <- levels(seurat)
names(colTec) <- unique(seurat$technique)
names(colSmp) <- unique(seurat$dataset)
names(colCond) <- unique(seurat$cond2)
names(colID) <- unique(seurat$ID)
names(colOrig) <- unique(seurat$origin)
names(colIso) <- unique(seurat$isolation)
names(colProc) <- unique(seurat$processing)
names(colLab) <- unique(seurat$label)

```


## vis data {.tabset}
### clusters
```{r vis cluster}
DimPlot(seurat, reduction = "umap", cols=colPal)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### label
```{r vis label}
DimPlot(seurat, reduction = "umap",  group.by = "label", cols=colLab)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### technique
```{r vis tecnique}

DimPlot(seurat, reduction = "umap", group.by = "technique", cols=colTec)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

```


### Sample
```{r vis sample, fig.height=4, fig.width=9}

DimPlot(seurat, reduction = "umap", group.by = "dataset", cols=colSmp)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### ID
```{r vis ID}

DimPlot(seurat, reduction = "umap", group.by = "ID", cols=colID)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### Origin
```{r vis origin}

DimPlot(seurat, reduction = "umap", group.by = "origin", cols=colOrig)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### isolation
```{r vis isolation}

DimPlot(seurat, reduction = "umap", group.by = "isolation", cols=colIso)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### cond
```{r vis cond, fig.height=4, fig.width=5}

DimPlot(seurat, reduction = "umap", group.by = "cond2", cols=colCond)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### processing
```{r vis proc}

DimPlot(seurat, reduction = "umap", group.by = "processing", cols=colProc)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```



## cw DE genes

```{r cw DE genes}

Idents(seurat) <- seurat$cond2
grpVec <- unique(seurat$label)

clustDE <- lapply(grpVec, function(grp){
  grpSub <- unique(seurat$label)[which(
    unique(seurat$label)==grp)]
  seuratSub <- subset(seurat, label == grpSub)
  DEgenes <-FindAllMarkers(seuratSub, only.pos=T, logfc.threshold = 0.1,
                           min.pct = 0.01)
  if(nrow(DEgenes)>1){
    DEgenes <- DEgenes %>% filter(p_val_adj<0.01) %>%
      mutate(group=paste0(grp, "_", cluster)) %>% 
      mutate(geneID=gsub(".*\\.", "",  gene)) %>% 
      filter(nchar(geneID)>1)
    }
})

names(clustDE) <- grpVec

clustDE_Dat <- data.frame(do.call("rbind", clustDE))

write.table(clustDE_Dat,
            file=paste0(basedir, 
                              "/data/humanHeartsPlusGraz_intPatients_", 
                              "merged_Subset_cwDEGenes.txt"),
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

```



### avg Heatmap top cwDE genes {.tabset}
```{r avgHeat cwDE, warning=FALSE, echo = FALSE}

## only cluster with > 1 DE genes
deCnt <- lapply(clustDE, nrow)
deCnt <- data.frame(cnt=do.call("rbind", deCnt)) %>%
  rownames_to_column(var = "cluster") %>% filter(cnt>1)

grpVec <- as.character(deCnt$cluster)
seurat$label_plus_cond2 <- paste0(seurat$label, "_", seurat$cond2)
seurat$label_plus_cond2 <- as.factor(seurat$label_plus_cond2)
Idents(seurat) <- seurat$label_plus_cond2

template_ah <- c(
    "#### {{grp}}\n",
    "```{r avgHeat cwDE {{grp}},echo = FALSE,fig.height=10,fig.width=10}\n",
    "DeSub <- clustDE[[as.character('{{grp}}')]]",
    "DeSel <- DeSub %>% top_n(., 30, avg_log2FC)",
    "gapVecCol <- seq(2, length(levels(seurat$label_plus_cond2)), by=2)",
    "pOut <- avgHeatmap(seurat = seurat, selGenes = DeSel,
                     colVecIdent = colLab, colVecCond=colCond,
                     ordVec=levels(seurat$label_plus_cond2),
                     gapVecR=NULL, gapVecC=gapVecCol,cc=FALSE,
                     cr=T, condCol=T)\n",
    "```\n",
    "\n"
)
    
plots_ah <- lapply(grpVec, 
  function(grp) knitr::knit_expand(text = template_ah)
)

```

`r knitr::knit(text = unlist(plots_ah))`


### GSEA clusterProfiler cw DEgenes 

```{r GSEA clusterprofiler, fig.height=6, fig.width=12}

clVec <- levels(seurat$label_plus_cond2)
GOcons <- lapply(clVec, function(cl){
  clustDE_DatSub <- clustDE_Dat[which(clustDE_Dat$group == cl),] %>% 
    mutate(ENS=gsub("\\..*$", "", gene)) #%>% 
    #slice_min(., max_pval, n=200)
  egoSS <- enrichGO(gene      = unique(clustDE_DatSub$ENS),
                OrgDb         = org.Hs.eg.db,
                keyType       = 'ENSEMBL',
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05)
  egoSS <- setReadable(egoSS, OrgDb = org.Hs.eg.db)
  egoSSres <- egoSS@result %>% filter(p.adjust < 0.05) %>% 
    mutate(subset=cl)
})

names(GOcons) <- clVec

## table to select pathways
GOconsDat <- do.call("rbind", GOcons)

write.table(GOconsDat, quote=F, row.names = T, col.names = T, sep= "\t",
            file = paste0(basedir,"/data/humanHeartsPlusGraz_intPatients_", 
                              "merged_Subset_cwDEGO.txt"))
	


```

### vis sel GO Fibroblasts
```{r vis sel GO Fibroblasts}

selGO <- read_tsv(paste0(basedir,"/data/GSEA/selGO_Fibroblasts.txt"))
GOconsDatSel <- GOconsDat %>% filter(ID %in% selGO$GOterm) %>% 
  mutate(cond2 = gsub(".*_", "", subset)) %>% 
  mutate(label= gsub("_.*", "", subset)) %>% 
  filter(label == "Fibroblast")

grpVec <- unique(selGO$cond2)
lapply(grpVec, function(grp){
  selGODat <- GOconsDatSel %>% filter(cond2 == grp)
  selGODat <- selGODat %>% mutate(qscore=-log(p.adjust, base=10)) 
  p <- ggbarplot(selGODat, x = "Description", y = "qscore",
          fill = "cond2",               
          color = "cond2",            
          palette = colCond,            
          sort.val = "asc",           
          sort.by.groups = TRUE      
          #x.text.angle = 90           
          ) + 
  rotate()
p
})


```


### vis sel GO T cells
```{r vis sel GO Tcell}

selGO <- read_tsv(paste0(basedir,"/data/GSEA/selGO_Tcell.txt"))
GOconsDatSel <- GOconsDat %>% filter(ID %in% selGO$GOterm) %>% 
  mutate(cond2 = gsub(".*_", "", subset)) %>% 
  mutate(label= gsub("_.*", "", subset)) %>% 
  filter(label == "Tcell")

grpVec <- unique(selGO$cond2)
lapply(grpVec, function(grp){
  selGODat <- GOconsDatSel %>% filter(cond2 == grp)
  selGODat <- selGODat %>% mutate(qscore=-log(p.adjust, base=10)) 
  p <- ggbarplot(selGODat, x = "Description", y = "qscore",
          fill = "cond2",               
          color = "cond2",            
          palette = colCond,            
          sort.val = "asc",           
          sort.by.groups = TRUE      
          #x.text.angle = 90           
          ) + 
  rotate()
p
})


```



## signatures viridis
### split by grp
```{r signature cut 1.5, fig.height=3.5, fig.width=5}

signDat <- read_delim(file = paste0(basedir,
                    "/data/SelSignaturesTreat2.txt"),
                    delim = "\t")
genes <- data.frame(geneID=rownames(seurat)) %>% 
  mutate(gene=gsub("^.*\\.", "", geneID))
signDat <- signDat %>% left_join(.,genes, by="gene")
allSign <- unique(signDat$signature)

DefaultAssay(object = seurat) <- "integrated"
sce2 <- as.SingleCellExperiment(seurat)

DefaultAssay(object = seurat) <- "RNA"
sce <- as.SingleCellExperiment(seurat)
reducedDims(sce) <- list(PCA=reducedDim(sce2, "PCA"),
                         TSNE=reducedDim(sce2, "TSNE"),
                         UMAP=reducedDim(sce2, "UMAP"))

treatGrps <- unique(sce$cond2)

cutOff <- 3
pal = viridis(100)
sc <- scale_colour_gradientn(colours = pal, limits=c(0, cutOff))

lapply(unique(signDat$signature), function(sign){
  signGenes <- signDat %>% dplyr::filter(signature == sign)
  sceSub <- sce[which(rownames(sce) %in% signGenes$geneID),]
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts)))/nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign[which(sceSub$sign > cutOff)] <- cutOff
  sceSub$sign[which(sceSub$sign < 0)] <- 0
  lapply(treatGrps, function(treat){
    sceSubT <- sceSub[, which(sceSub$cond2 == treat)]
    p <- visGroup_adapt(sceSubT, 'sign', dim_red = 'UMAP') +
    sc +
    guides(colour = guide_colourbar(title = '')) +
    ggtitle(paste0(sign, ' signature - ', treat)) +
    theme_classic() + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x='Dimension 1', y='Dimension 2')
    p
  })
})


cutOff <- 2
pal = viridis(100)
sc <- scale_colour_gradientn(colours = pal, limits=c(0, cutOff))

lapply(unique(signDat$signature), function(sign){
  signGenes <- signDat %>% dplyr::filter(signature == sign)
  sceSub <- sce[which(rownames(sce) %in% signGenes$geneID),]
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts)))/nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign[which(sceSub$sign > cutOff)] <- cutOff
  sceSub$sign[which(sceSub$sign < 0)] <- 0
  lapply(treatGrps, function(treat){
    sceSubT <- sceSub[, which(sceSub$cond2 == treat)]
    p <- visGroup_adapt(sceSubT, 'sign', dim_red = 'UMAP') +
    sc +
    guides(colour = guide_colourbar(title = '')) +
    ggtitle(paste0(sign, ' signature - ', treat)) +
    theme_classic() + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x='Dimension 1', y='Dimension 2')
    p
  })
})

cutOff <- 1.5
pal = viridis(100)
sc <- scale_colour_gradientn(colours = pal, limits=c(0, cutOff))

lapply(unique(signDat$signature), function(sign){
  signGenes <- signDat %>% dplyr::filter(signature == sign)
  sceSub <- sce[which(rownames(sce) %in% signGenes$geneID),]
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts)))/nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign[which(sceSub$sign > cutOff)] <- cutOff
  sceSub$sign[which(sceSub$sign < 0)] <- 0
  lapply(treatGrps, function(treat){
    sceSubT <- sceSub[, which(sceSub$cond2 == treat)]
    p <- visGroup_adapt(sceSubT, 'sign', dim_red = 'UMAP') +
    sc +
    guides(colour = guide_colourbar(title = '')) +
    ggtitle(paste0(sign, ' signature - ', treat)) +
    theme_classic() + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x='Dimension 1', y='Dimension 2')
    p
  })
})

```

## signatures red blue
### split by grp
```{r signature rdbl, fig.height=3.5, fig.width=5}


cutOff <- 3
pal = colorRampPalette(c("#053061", "#f7f7f7","#85122d"))(100)
sc <- scale_colour_gradientn(colours = pal, limits=c(0, cutOff))

lapply(unique(signDat$signature), function(sign){
  signGenes <- signDat %>% dplyr::filter(signature == sign)
  sceSub <- sce[which(rownames(sce) %in% signGenes$geneID),]
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts)))/nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign[which(sceSub$sign > cutOff)] <- cutOff
  sceSub$sign[which(sceSub$sign < 0)] <- 0
  lapply(treatGrps, function(treat){
    sceSubT <- sceSub[, which(sceSub$cond2 == treat)]
    p <- visGroup_adapt(sceSubT, 'sign', dim_red = 'UMAP') +
    sc +
    guides(colour = guide_colourbar(title = '')) +
    ggtitle(paste0(sign, ' signature - ', treat)) +
    theme_classic() + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x='Dimension 1', y='Dimension 2')
    p
  })
})


cutOff <- 2
pal = colorRampPalette(c("#053061", "#f7f7f7","#85122d"))(100)
sc <- scale_colour_gradientn(colours = pal, limits=c(0, cutOff))

lapply(unique(signDat$signature), function(sign){
  signGenes <- signDat %>% dplyr::filter(signature == sign)
  sceSub <- sce[which(rownames(sce) %in% signGenes$geneID),]
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts)))/nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign[which(sceSub$sign > cutOff)] <- cutOff
  sceSub$sign[which(sceSub$sign < 0)] <- 0
  lapply(treatGrps, function(treat){
    sceSubT <- sceSub[, which(sceSub$cond2 == treat)]
    p <- visGroup_adapt(sceSubT, 'sign', dim_red = 'UMAP') +
    sc +
    guides(colour = guide_colourbar(title = '')) +
    ggtitle(paste0(sign, ' signature - ', treat)) +
    theme_classic() + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x='Dimension 1', y='Dimension 2')
    p
  })
})

cutOff <- 1.5
pal = colorRampPalette(c("#053061", "#f7f7f7","#85122d"))(100)
sc <- scale_colour_gradientn(colours = pal, limits=c(0, cutOff))

lapply(unique(signDat$signature), function(sign){
  signGenes <- signDat %>% dplyr::filter(signature == sign)
  sceSub <- sce[which(rownames(sce) %in% signGenes$geneID),]
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts)))/nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign[which(sceSub$sign > cutOff)] <- cutOff
  sceSub$sign[which(sceSub$sign < 0)] <- 0
  lapply(treatGrps, function(treat){
    sceSubT <- sceSub[, which(sceSub$cond2 == treat)]
    p <- visGroup_adapt(sceSubT, 'sign', dim_red = 'UMAP') +
    sc +
    guides(colour = guide_colourbar(title = '')) +
    ggtitle(paste0(sign, ' signature - ', treat)) +
    theme_classic() + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x='Dimension 1', y='Dimension 2')
    p
  })
})

```




## session info
```{r session info}
sessionInfo()
date()
```



