---
title: "Merge samples"
author: "Mechthild Lütge"
date: "14 May 2020"
output:
  html_document:
    self_contained: no
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
set.seed(1000)
```

## load packages
```{r libs}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(tidyverse)
  library(Seurat)
  library(magrittr)
  library(dplyr)
  library(purrr)
  library(ggplot2)
  library(here)
  library(runSeurat3)
  library(ggsci)
  library(ggpubr)
  library(pheatmap)
})

```


```{r avg heatmap function, eval=T, include=FALSE}

avgHeatmap <- function(seurat, selGenes, colVecIdent, colVecCond=NULL,
                       ordVec=NULL, gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                       cr=FALSE, condCol=FALSE){
  
  ## format gene names (depends on how gene list is inputed)
  selGenes <- selGenes %>% 
      mutate(labelNam = (str_split(gene, '\\.', simplify = T)[,2]))
  selGenes <- selGenes$labelNam
  
  ## assay data
  clusterAssigned <- as.data.frame(Idents(seurat)) %>%
  dplyr::mutate(cell=rownames(.))
  colnames(clusterAssigned)[1] <- "ident"
  seuratDat <- GetAssayData(seurat)
  
  ## genes of interest
  genes <- data.frame(gene=rownames(seurat)) %>% 
    mutate(geneID=gsub("^.*\\.", "", gene)) %>% filter(geneID %in% selGenes)

  ## matrix with averaged cnts per ident
  logNormExpres <- as.data.frame(t(as.matrix(
    seuratDat[which(rownames(seuratDat) %in% genes$gene),])))
  logNormExpres <- logNormExpres %>% dplyr::mutate(cell=rownames(.)) %>%
    dplyr::left_join(.,clusterAssigned, by=c("cell")) %>%
    dplyr::select(-cell) %>% dplyr::group_by(ident) %>%
    dplyr::summarise_all(mean)
  logNormExpresMa <- logNormExpres %>% dplyr::select(-ident) %>% as.matrix()
  rownames(logNormExpresMa) <- logNormExpres$ident
  logNormExpresMa <- t(logNormExpresMa)
  rownames(logNormExpresMa) <- gsub("^.*?\\.","",rownames(logNormExpresMa))
  
  ## remove genes if they are all the same in all groups
  ind <- apply(logNormExpresMa, 1, sd) == 0
  logNormExpresMa <- logNormExpresMa[!ind,]
  genes <- genes[!ind,]

  ## color columns according to cluster
  annotation_col <- as.data.frame(gsub("(^.*?_)","",
                                       colnames(logNormExpresMa)))%>%
    dplyr::mutate(celltype=gsub("(_.*$)","",colnames(logNormExpresMa)))
  colnames(annotation_col)[1] <- "col1"
  annotation_col <- annotation_col %>%
    dplyr::mutate(cond = gsub("(^[0-9]_?)","",col1)) %>%
    dplyr::select(cond, celltype)
  rownames(annotation_col) <- colnames(logNormExpresMa) 

  ann_colors = list(
      cond = colVecCond,
      celltype=colVecIdent)
  if(is.null(ann_colors$cond)){
    annotation_col$cond <- NULL
  }
  
  ## adjust order
  logNormExpresMa <- logNormExpresMa[selGenes,]
  if(is.null(ordVec)){
    ordVec <- levels(seurat)
  }
  logNormExpresMa <- logNormExpresMa[,ordVec]

  ## scaled row-wise
  pheatmap(logNormExpresMa, scale="row" ,treeheight_row = 0, cluster_rows = cr, 
         cluster_cols = cc,
         color = colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50),
         annotation_col = annotation_col, cellwidth=15, cellheight=10,
         annotation_colors = ann_colors, gaps_row = gapVecR, gaps_col = gapVecC)
}

```


## set dir
```{r set dir}
basedir <- here()
seurat <- readRDS(file = paste0(basedir, 
                              "/data/humanHearts_merged_seurat.rds"))
seurat$ID[which(seurat$ID == "ID22_24")] <- "ID2224"
seurat$ID[which(seurat$ID == "ID23_25")] <- "ID2325"
Idents(seurat) <- seurat$RNA_snn_res.0.25

```



## color vectors
```{r colVec}

colPal <- pal_igv()(length(levels(seurat)))
colTec <- pal_jama()(length(unique(seurat$technique)))
colSmp <- c(pal_uchicago()(8), pal_npg()(8), pal_aaas()(10))[1:length(unique(seurat$dataset))]
colLoc <- pal_npg()(length(unique(seurat$location)))
colBatch <- c(pal_jco()(10), pal_npg()(10))[1:length(unique(seurat$ID))]
colOrig <- pal_futurama()(length(unique(seurat$origin)))
colIso <- pal_nejm()(length(unique(seurat$isolation)))

names(colPal) <- levels(seurat)
names(colTec) <- unique(seurat$technique)
names(colSmp) <- unique(seurat$dataset)
names(colLoc) <- unique(seurat$location)
names(colBatch) <- unique(seurat$ID)
names(colOrig) <- unique(seurat$origin)
names(colIso) <- unique(seurat$isolation)

```


## vis data {.tabset}
### clusters
```{r vis cluster}
DimPlot(seurat, reduction = "umap", cols=colPal)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### technique
```{r vis tecnique}

DimPlot(seurat, reduction = "umap", group.by = "technique", cols=colTec)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

```


### Sample
```{r vis sample, fig.height=4, fig.width=8}

DimPlot(seurat, reduction = "umap", group.by = "dataset", cols=colSmp)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### ID
```{r vis ID}

DimPlot(seurat, reduction = "umap", group.by = "ID", cols=colBatch)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### Origin
```{r vis origin}

DimPlot(seurat, reduction = "umap", group.by = "origin", cols=colOrig)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### isolation
```{r vis isolation}

DimPlot(seurat, reduction = "umap", group.by = "isolation", cols=colIso)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### location
```{r vis location}

DimPlot(seurat, reduction = "umap", group.by = "location", cols=colLoc)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```



## DE genes EMB 29 vs other EMB
```{r DE genes EMB 29}

seuratSub <- subset(seurat, origin != "ECMO")
seuratSub$grp <- "other"
seuratSub$grp[which(seuratSub$ID == "ID29")] <- "EMB29"
Idents(seuratSub) <- seuratSub$grp

DEgenes <- FindAllMarkers(object = seuratSub, assay ="RNA",
                                     only.pos = TRUE, min.pct = 0.25,
                                     logfc.threshold = 0.25,
                                     test.use = "wilcox")

```


### top 20 marker genes per grp

```{r top DE genes EMB29, fig.height=12, fig.width=8}

cluster <- levels(seurat)
selGenesAll <- DEgenes %>% group_by(cluster) %>% 
  top_n(-20, p_val_adj) %>% 
  top_n(20, avg_log2FC)
selGenesAll <- selGenesAll %>% mutate(geneIDval=gsub("^.*\\.", "", gene)) %>% filter(nchar(geneIDval)>1)

Idents(seuratSub) <- seuratSub$ID
pOut <- avgHeatmap(seurat = seuratSub, selGenes = selGenesAll,
                  colVecIdent = colBatch, 
                  ordVec=levels(seuratSub),
                  gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                  cr=T, condCol=F)

Idents(seuratSub) <- seuratSub$RNA_snn_res.0.25
pOut <- avgHeatmap(seurat = seuratSub, selGenes = selGenesAll,
                  colVecIdent = colPal, 
                  ordVec=levels(seuratSub),
                  gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                  cr=T, condCol=F)

```

## remove EMB29
```{r remove EMB29}
seurat <- subset(seurat, ID != "ID29")
Idents(seurat) <- seurat$RNA_snn_res.0.25
```


## vis data without EMB29 {.tabset}
### clusters
```{r vis cluster without EMB29}
DimPlot(seurat, reduction = "umap", cols=colPal)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### technique
```{r vis tecnique without EMB29}

DimPlot(seurat, reduction = "umap", group.by = "technique", cols=colTec)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")

```


### Sample
```{r vis sample without EMB29, fig.height=4, fig.width=8}

DimPlot(seurat, reduction = "umap", group.by = "dataset", cols=colSmp)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```

### ID
```{r vis ID without EMB29}

DimPlot(seurat, reduction = "umap", group.by = "ID", cols=colBatch)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### Origin
```{r vis origin without EMB29}

DimPlot(seurat, reduction = "umap", group.by = "origin", cols=colOrig)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### isolation
```{r vis isolation without EMB29}

DimPlot(seurat, reduction = "umap", group.by = "isolation", cols=colIso)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```


### location
```{r vis location without EMB29}

DimPlot(seurat, reduction = "umap", group.by = "location", cols=colLoc)+
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlab("UMAP1") +
  ylab("UMAP2")


```



## overall DE genes EMB 31 vs other EMB without EMB29
```{r DE genes EMB 31 without EMB29}

seuratSub <- subset(seurat, origin != "ECMO")
seuratSub$grp <- "other"
seuratSub$grp[which(seuratSub$ID == "ID31")] <- "EMB31"
Idents(seuratSub) <- seuratSub$grp

DEgenes31 <- FindAllMarkers(object = seuratSub, assay ="RNA",
                                     only.pos = TRUE, min.pct = 0.25,
                                     logfc.threshold = 0.25,
                                     test.use = "wilcox")

```


### top 20 marker genes per grp

```{r top DE genes EMB31 without EMB29, fig.height=12, fig.width=8}

cluster <- levels(seurat)
selGenesAll <- DEgenes31 %>% group_by(cluster) %>% 
  top_n(-20, p_val_adj) %>% 
  top_n(20, avg_log2FC)
selGenesAll <- selGenesAll %>% mutate(geneIDval=gsub("^.*\\.", "", gene)) %>% filter(nchar(geneIDval)>1)

Idents(seuratSub) <- seuratSub$ID
pOut <- avgHeatmap(seurat = seuratSub, selGenes = selGenesAll,
                  colVecIdent = colBatch, 
                  ordVec=levels(seuratSub),
                  gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                  cr=T, condCol=F)

Idents(seuratSub) <- seuratSub$RNA_snn_res.0.25
pOut <- avgHeatmap(seurat = seuratSub, selGenes = selGenesAll,
                  colVecIdent = colPal, 
                  ordVec=levels(seuratSub),
                  gapVecR=NULL, gapVecC=NULL,cc=FALSE,
                  cr=T, condCol=F)

```


## clusterwise DE genes EMB31 without EMB29

```{r cw DE genes EMB31}

Idents(seuratSub) <- seuratSub$grp
grpVec <- unique(seuratSub$RNA_snn_res.0.25)

clustDE <- lapply(grpVec, function(grp){
  grpSub <- unique(seuratSub$RNA_snn_res.0.25)[which(
    unique(seuratSub$RNA_snn_res.0.25)==grp)]
  seuratSub2 <- subset(seuratSub, RNA_snn_res.0.25 == grpSub)
  DEgenes <-FindAllMarkers(seuratSub2, only.pos=T, logfc.threshold = 0.25,
                           min.pct = 0.1)
  if(nrow(DEgenes)>1){
    DEgenes <- DEgenes %>% filter(p_val_adj<0.01) %>%
      mutate(group=paste0(grp, "_", cluster)) %>% 
      mutate(geneID=gsub(".*\\.", "",  gene)) %>% 
      filter(nchar(gene)>1)
    }
})

names(clustDE) <- grpVec

clustDE_Dat <- data.frame(do.call("rbind", clustDE))
```



### avg Heatmap top cwDE genes {.tabset}
```{r avgHeat cwDE, warning=FALSE, echo = FALSE}

colGrp <- pal_jama()(length(unique(seuratSub$grp)))
names(colGrp) <- unique(seuratSub$grp)
## only cluster with > 1 DE genes
deCnt <- lapply(clustDE, nrow)
deCnt <- data.frame(cnt=do.call("rbind", deCnt)) %>%
  rownames_to_column(var = "cluster") %>% filter(cnt>1)

grpVec <- as.character(deCnt$cluster)
seuratSub$clust_plus_grp <- as.factor(paste0(seuratSub$RNA_snn_res.0.25, "_",
                                             seuratSub$grp))
Idents(seuratSub) <- seuratSub$clust_plus_grp

template_ah <- c(
    "#### {{grp}}\n",
    "```{r avgHeat cwDE {{grp}},echo = FALSE,fig.height=10,fig.width=10}\n",
    "DeSub <- clustDE[[as.character('{{grp}}')]]",
    "DeSel <- DeSub %>% top_n(., 30, avg_log2FC) %>% 
      filter(nchar(geneID) >1)",
    "gapVecCol <- seq(2, length(levels(seuratSub$clust_plus_grp)), by=2)",
    "pOut <- avgHeatmap(seurat = seuratSub, selGenes = DeSel,
                     colVecIdent = colPal, colVecCond=colGrp,
                     ordVec=levels(seuratSub$clust_plus_grp),
                     gapVecR=NULL, gapVecC=gapVecCol,cc=FALSE,
                     cr=T, condCol=T)\n",
    "```\n",
    "\n"
)
    
plots_ah <- lapply(grpVec, 
  function(grp) knitr::knit_expand(text = template_ah)
)

```

`r knitr::knit(text = unlist(plots_ah))`



## save objects
```{r save}
Idents(seurat) <- Idents(seurat) <- seurat$RNA_snn_res.0.25
saveRDS(seurat, file = paste0(basedir, 
                              "/data/humanHearts_merged_woEMB29_seurat.rds"))

write.table(DEgenes,
            file=paste0(basedir,
                        "/data/humanHearts_EMB29vsOther_overallDEGenes.txt"),
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

write.table(DEgenes31,
            file=paste0(basedir,
                        "/data/humanHearts_EMB31vsOther_overallDEGenes.txt"),
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

write.table(clustDE_Dat,
            file=paste0(basedir,
                        "/data/humanHearts_EMB29vsOther_cwDEGenes.txt"),
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

```


## session info
```{r session info}
sessionInfo()
date()
```




